MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-13 main/MSstatsPTM_Main.tex
Updated shigella to time series model

\subsection*{Computer simulations}
\label{sec:comp_sim_procedure}

\paragraph*{Dataset 1 : Computer simulation 1 - label free}

@@

\paragraph*{Dataset 2 : Computer simulation 2 - missing values and low replicates}

@@

\subsection*{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}

@@

\subsection*{Dataset 4 : Human-1mix-TMT - ubiquitination}

@@

\subsection*{Dataset 5 : Mouse-2mix-TMT - phosphorylation}

@@

\subsection*{Dataset 6 : Human label free quantification - no global profiling run}

@@

\subsubsection*{Dataset 1 : Computer simulation 1 - label free}

@@

\subsubsection*{Dataset 2 : Computer simulation 2 - missing values and low replicates}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Dataset 4 : Human-1mix-TMT - ubiquitination}

The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}. In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides become insignificant after adjustment than become significant. 3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are correct in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.

@@

\subsubsection*{Dataset 5 : Mouse-2mix-TMT - phosphorylation}

The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}. In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides become insignificant, 4,947 became significant, while 39,550 peptides are significant in both models. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 776 peptides became insignificant due to an increase in variance. This is still a very small portion (4.02\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

In Figure \ref{fig:shigella_profile} the profile plot of two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.  The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run shows a similar log fold change of $2.015$ between the same conditions. This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.275$. In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.212$ between conditions WT\_Early and WT\_Uninfected. This is contrast with a negative fold change in the global profiling run of $-.616$. In adjusting the modified peptide for changes in the global protein, the log fold change becomes .828 and the adjusted pvalue goes from $.452$ to $.037$.

\subsubsection*{Dataset 6 : Human label free quantification - no global profiling run}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{Dataset 1 : Computer simulation 1 - label free. All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
}\label{fig:fdr_all_models}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{Dataset 2 : Computer simulation 2 - missing values and low replicates. The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values. In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. The lowest performaning method was $t$-test. Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
\label{fig:sim_accuracy}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
\caption{Dataset 2 : Computer simulation 2 - missing values and low replicates. In the second computer simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75. The proposed method in this simulation had a visibly tighter distribution around the median when compared to Limma and $t$-test. While the mean estimated fold changes were similar, the proposed method exhibited a stronger performance in correctly estimating the fold change for all peptides.}
\label{fig:sim_boxplot}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
\caption{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched. Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment. After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation. Additionally the background grey colored peptides showed many false positives before adjustment. After adjustment the false positives were decreased considerably.
}\label{fig:spikein_prop_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
\caption{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched. Here the Limma method was used to model the spike in experiment. Using Limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level. However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue. In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
}\label{fig:spikein_limma_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{Dataset 4 : Human-1mix-TMT - ubiquitination. The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment. More PTMs became insignificant after adjustment then became significant. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.}
\label{fig:ipah_venn_diagramm}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{Dataset 4 : Human-1mix-TMT - ubiquitination. Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$. The individual PSM features are shown in grey, while the feature summarization is shown in red. When looking at the summary of the modification and global protein it is clear the conditions follow different trends. Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot. This indicates the modification is confounded with changes in the unmodified protein.}
\label{fig:ipah_profile}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=.65\textwidth]{images/shig_venn_diagramm.png}
\caption{Dataset 5 : Mouse-2mix-TMT - phosphorylation. The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment. Again more PTMs became insignificant after adjustment then became significant.}
\label{fig:shig_venn_d}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.95\textwidth]{images/No_Difference_Shigella_Profile_Plot}
	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$. The individual PSM features are shown in grey, while the feature summarization is shown in red. When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend. Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run. This indicates the movement is driven by changes in global protein.}
 \end{subfigure}\vspace{5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.95\textwidth]{images/Difference_Shigella_Profile_Plot}
	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$. In this case the modification and global protein trend in different directions. Specifically, comparing WT\_Uninfect and WT\_Early there is a slightly positive change in abundance, however in the global profiling there was a negative change. In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance. Additionally this profile plot shows the large difference in available features between modifications and global protein.}
 \end{subfigure}
\caption{Dataset 5 : Mouse-2mix-TMT - phosphorylation. Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
\label{fig:shigella_profile}
\end{figure}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm}
	\caption{The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment. Here many more PTMs became insignificant then became significant after adjustment. This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.}
 \end{subfigure}\vspace{-5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm_matching_only}
	\caption{Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed. In this case we see significantly less peptides become insignificant after adjustment. This highlights the need for a global profiling run if protein adjustment is going to performed.}
 \end{subfigure}
 \caption{Dataset 6 : Human label free quantification - no global profiling run. Overlap of significant PTMs before and after protein adjustment.}
\label{fig:usp30_ven_d}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/USP30_profile_plot.png}
\caption{Dataset 6 : Human label free quantification - no global profiling run. Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$. The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE. This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
\label{fig:usp30_profile}
\end{figure}

>>>>

\subsection*{Dataset 1 : Computer simulation 1 - Label-free} 
\label{sec:comp_sim_procedure1}

@@

\subsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates} 
\label{sec:comp_sim_procedure2}

@@

\subsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}

@@

\subsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}

@@

\subsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}

@@

\subsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}

@@

\subsubsection*{Dataset 1 : Computer simulation 1 - Label-free}

@@

\subsubsection*{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Dataset 4 : Human - Ubiquitination - 1mix-TMT}

The results of this experiment are summarized in Figures \ref{fig:ipah_venn_diagramm} and \ref{fig:ipah_profile}. The experiment was modeled as a group comparison. In Figure \ref{fig:ipah_venn_diagramm} the number of significant modified peptides before and after adjustment is shown. We can see that more peptides become insignificant after adjustment than become significant. 3,236 peptides become insignificant, 1352 peptides become significant, while 4,282 peptides are significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues are correct in adjustment, and the true abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was truly due to changes in global protein abundance.

@@

\subsubsection*{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series}

The results of this experiment are summarized in Figures \ref{fig:shig_venn_d} and \ref{fig:shigella_profile}. The experiment was modeled as a time series, with the same bioreplicate being measured at each time point. In Figure \ref{fig:shig_venn_d} the number of significant modified peptides before and after adjustment is shown. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides become insignificant, 4,947 became significant, while 39,550 peptides are significant in both models. Again we tested if the decrease in significant peptides was due to the increased variance that comes from adjustment, or if it was mainly due to removing convolution with change in the umodified protein. This was tested by looking for modified peptides who's adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied on this experiment, 548 peptides became insignificant due to an increase in variance. This is still a very small portion (3.4\%) of the peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

In Figure \ref{fig:shigella_profile} the profile plot of two specific modified peptides, $TTP\_MOUSE|P22893\_S178$ and $KI67\_MOUSE|E9PVX6\_T215$, show the power of the proposed method.  The profile plot of modified peptide $TTP\_MOUSE|P22893\_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run shows a similar log fold change of $2.014$ between the same conditions. This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.248$. In contrast $KI67\_MOUSE|E9PVX6\_T215$ originally shows a small positive log fold change of $.185$ between conditions WT\_Early and WT\_Uninfected. This is contrast with a negative fold change in the global profiling run of $-.619$. In adjusting the modified peptide for changes in the global protein, the log fold change becomes $.804$ and the adjusted pvalue goes from $.338$ to $.0156$.

\subsubsection*{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{Dataset 1 : Computer simulation 1 - Label-free. All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
}\label{fig:fdr_all_models}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates. The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values. In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. The lowest performaning method was $t$-test. Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
\label{fig:sim_accuracy}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
\caption{Dataset 2 : Computer simulation 2 - Label-free missing values and low replicates. In the second computer simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75. The proposed method in this simulation had a visibly tighter distribution around the median when compared to Limma and $t$-test. While the mean estimated fold changes were similar, the proposed method exhibited a stronger performance in correctly estimating the fold change for all peptides.}
\label{fig:sim_boxplot}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free. Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment. After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation. Additionally the background grey colored peptides showed many false positives before adjustment. After adjustment the false positives were decreased considerably.
}\label{fig:spikein_prop_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
\caption{Dataset 3 : SpikeIn benchmark - Ubiquitination - Label-free. Here the Limma method was used to model the spike in experiment. Using Limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level. However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue. In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
}\label{fig:spikein_limma_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[height=.65\textwidth]{images/ipah_venn_diagramm.png}
\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT. The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment. More PTMs became insignificant after adjustment then became significant. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.}
\label{fig:ipah_venn_diagramm}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{Dataset 4 : Human - Ubiquitination - 1mix-TMT. Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$. The individual PSM features are shown in grey, while the feature summarization is shown in red. When looking at the summary of the modification and global protein it is clear the conditions follow different trends. Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot. This indicates the modification is confounded with changes in the unmodified protein.}
\label{fig:ipah_profile}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=.65\textwidth]{images/shig_venn_diagramm.png}
\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series. The overlap of differencial modified peptides between the PTM model with and without global protein level adjustment. Again more PTMs became insignificant after adjustment then became significant.}
\label{fig:shig_venn_d}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.95\textwidth]{images/No_Difference_Shigella_Profile_Plot}
	\caption{Comparing the global profiling of protein $TTP\_MOUSE|P22893$ with the modification of the protein at site $S178$. The individual PSM features are shown in grey, while the feature summarization is shown in red. When looking at the summary of the modification and global protein it is clear the difference between conditions follow the same trend. Specifically, there is a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run. This indicates the movement is driven by changes in global protein.}
 \end{subfigure}\vspace{5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.95\textwidth]{images/Difference_Shigella_Profile_Plot}
	\caption{Comparing the global profiling of protein $KI67\_MOUSE|E9PVX6$ with the modification of the protein at site $T215$. In this case the modification and global protein trend in different directions. Specifically, comparing WT\_Uninfect and WT\_Early there is a slightly positive change in abundance, however in the global profiling there was a negative change. In this case the profile plot indicates the effect of the modification is masked by the change in global protein abundance. Additionally this profile plot shows the large difference in available features between modifications and global protein.}
 \end{subfigure}
\caption{Dataset 5 : Mouse - Phosphorylation - 2mix-TMT time series. Profile plots for PTMs P22893\_S178 and E9PVX6\_T215 in the Shigella flexneri experiment.}
\label{fig:shigella_profile}
\end{figure}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm}
	\caption{The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment. Here many more PTMs became insignificant then became significant after adjustment. This is due to not having a global profiling run, resulting in a large lack of overlap between the modified peptides and unmodified proteins.}
 \end{subfigure}\vspace{-5mm}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[height=.5\textwidth]{images/usp30_venn_diagramm_matching_only}
	\caption{Here we make the same comparison but only for modified peptides with a matching unmodified protein, so adjustment can be performed. In this case we see significantly less peptides become insignificant after adjustment. This highlights the need for a global profiling run if protein adjustment is going to performed.}
 \end{subfigure}
 \caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run. Overlap of significant PTMs before and after protein adjustment.}
\label{fig:usp30_ven_d}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/USP30_profile_plot.png}
\caption{Dataset 6 : Human - Ubiquitination - Label-free no global profiling run. Comparing the global profiling of protein $P52209$ with the modification of the protein at site $K059$. The modification appears generally unchanged between conditions, whereas the global profiling run shows the CCCP and Combo conditions at a higher relative abundance compared to the Control and USP30\_OE. This indicates that the modification actually had a major effect when comparing CCCP and Combo to Control and USP30\_OE, which would have been missed without adjusting for global protein changes.}
\label{fig:usp30_profile}
\end{figure}
